---
title: "PHMM"
author: "Ben Halliwell & Luke Yates"
date: "11/06/2021"
bibliography: bib_phmm.bib
link-citations: true
output: 
  html_document:
    number_sections: false
    toc: true 
    toc_float:
      collapsed: false
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse); library(ggplot2); library(knitr); library(brms)
library(rstan); library(geiger)
knitr::opts_chunk$set(echo = TRUE, dev="png",
               dpi=96)
```

# Tutorial

## Introduction


### Tree

```{r}
t.toy <- sim.bdtree(b=1, d=0, stop="taxa", n=100, extinct=FALSE) # may be best not to use pure birth trees (Adams and Collyer 2018)
plot(t.toy)

```


### Model 1 - univariate Gaussian

```{r}
1+1

```

#### model explanation

text + maths

\begin{eqnarray}
y_i &=& \beta_0 + b_i + \epsilon_i\\[2mm]
b_i &\sim& \mbox{MVN}(0, \sigma_pA)\\
\begin{pmatrix}
1 \\ a 
\end{pmatrix}&\sim&
\end{eqnarray}

#### simulate data under model 1





#### sub-heading

The  ...

For maths use dollars signs to get latex commands: $\lambda$

For a reference use at symbol @Royle2004 , see the bib_phmm.bib file for standard bibtex. 

Example equation:
\begin{eqnarray}
n_{ij}|N_{ij} &\sim& \mbox{Bin}(N_{ij},p_{ij})\\
N_{ij} &\sim& f(\theta)
\end{eqnarray}

R code:

```{r}
1
```


# References

