---
title: "PHMM"
author: "Ben Halliwell, Luke Yates & Barbara Holland"
header-includes:
  - \usepackage{mathtools}
date: "11/06/2021"
bibliography: bib_phmm.bib
link-citations: true
output: 
  html_document:
    number_sections: false
    toc: true 
    toc_float:
      collapsed: false
      smooth_scroll: false
---

### N.B.
Pull before you start. After changes, commit and push.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse); library(ggplot2); library(knitr); library(brms)
library(rstan); library(geiger); library(ape)
knitr::opts_chunk$set(echo = TRUE, dev="png",
               dpi=96)
```

# Tutorial
\

### Introduction

Blah Blah

Brief explanation of Bayesian concepts and methods. Why they should be prefered for these sorts of models.

5 taxa tree example for exposition

Also introduce 50 taxa tree that we will use for all sims below 


\

### Model 1 - univariate Gaussian
\

#### Model Explanation

Our response variable, $y_i$, is modeled as a linear combination of the trait 
intercept, $\beta_0$, a phylogenetic random effect, $b_i$, and residual variance, $\epsilon_i$.


\begin{eqnarray}
\mathbf{y} &=& \boldsymbol{\mu} + \mathbf{b} + \mathbf{e}\\[2mm]
\end{eqnarray}\


If we look at the distribution of our random effect, $b$, it is multivariate normally distributed, with mean 0 and (co)variance given by $\sigma_{phy}A$, 

\

\begin{eqnarray}
b &\sim& \mbox{MVN}(0, \sigma_{phy}A)\\
\end{eqnarray}\

The phylogenetic variance (e.g., Brownian rate), $\sigma_{phy}$, is a scalar to be estimated by the model and $A$ is a fixed $n\times n$
phylogenetic VCV matrix, where $n$ is the number of taxa in the tree. This model therefore takes as input a phylogenetic tree (topology and branch lengths), which is assumed to be known without error. The $A$ matrix itself is derived by taking the inverse of the distance matrix of the tree. It gives the expected (co)variance among taxa in the value of $y$ assuming some model of evolution (e.g. BM) along the branches of that tree.

\

\begin{eqnarray}
A &=& 
\begin{pmatrix}A_{11} & A_{12} & \ldots & A_{1n} \\ 
                A_{21} & A_{22} & \ldots & A_{2n} \\
                \vdots & \vdots & \ddots & \vdots \\ 
                A_{n1} & A_{n2} & \ldots & A_{nn} \\ 
\end{pmatrix}
\end{eqnarray}\




The residual error may be represented similarly, substituting $A$ with the identity matrix, $I$ (i.e., an $n\times n$ matrix with 1 in all diagonal elements and 0 in all off-diagonal elements):

\

\begin{eqnarray}
\epsilon &\sim& \mbox{MVN}(0, \sigma_{res}I)\\
\end{eqnarray}\



Implementing these models in R is fairly straight forward in both MCMCglmm and brms. But first, let's simulate some data.
To simulate data under this model, we simply construct our variable, $y$, from the linear combination of predictors specified by
the model, by supplying vectors of length $n$ for each of the quantities $\beta_0$, $b$, and $\epsilon$. For Gaussian variables,
this can be done directly. For non-Gaussian variables we must make predictions on the link scale (see below).

\

```{r eval=FALSE}

## SHOW SIMULATION


## MCMCglmm 
p <- list(G = list(G1 = list(V = 1, nu = 0.002)), 
          R = list(V = 1, nu = 0.002))
m.1 <- MCMCglmm(y1 ~ 1,
                random = ~animal,
                pedigree=t.toy, 
                family = c("gaussian"), 
                data = d.toy, prior = p,
                nitt=210000, burnin=10000, thin=200,
                pr=TRUE,verbose = FALSE)

## BRMS
b.1 <- brm(y1 ~ 1 + (1|gr(animal, cov = A)), 
            data = d.toy, 
            family = gaussian(), 
            data2 = list(A = A.mat),
            cores=4,chains=4, 
            iter = 6000, thin = 3
)

```

\

Explanation and comparison of the syntax for MCMCglmm and brms 

Compare and contrast with PGLS (equivalent for the univariate Gaussian case?)

\

### Model 2 - multivariate Gaussian
\

#### Model Explanation

In a multivariate (MV) implementation, our response variables are modelled jointly, allowing both phylogenetic and residual (co)variances to be estimated. With this design, the phylogenetic random effects must also be realised jointly as a multivariate normal, with variance given by $\Omega$, the Kronecker product of a trait-level correlation matrix $\Sigma^{\mathrm{\tiny phy}}$ and the phylogenetic VCV matrix, $A$. With $m$ traits (response variables) and $n$ taxa in the phylogeny, $\Omega$ is an $mn\times mn$ VCV matrix containing phylogenetic (co)variances for all traits as well as between all traits. Notice that the two taxon-level variance structures we encountered in the univariate case above ($A$ and $I$) remain, but we must now also consider trait-level covariance. This is achieved via the Kronecker operation, e.g. $\Omega = \Sigma^{\mathrm{\tiny phy}}\otimes A$. A MV model with two Gaussian response variables takes the form,




\


\begin{eqnarray}
\begin{pmatrix}\mathbf{y_1} \\
\mathbf{y_2} \end{pmatrix} &=& 
\begin{pmatrix}\boldsymbol{\mu}_1 + \mathbf{b}_1 + \mathbf{e}_1 \\
               \boldsymbol{\mu}_2 + \mathbf{b}_2 + \mathbf{e}_2
\end{pmatrix}\\
\end{eqnarray}

\

where fixed effects are expressed as a linear combination of predictors,

\

\begin{eqnarray}
\begin{pmatrix}\boldsymbol{\mu_1} \\ \boldsymbol{\mu_2} \end{pmatrix} &=& 
\begin{pmatrix}\beta_{0,1}\mathbf{1} + \beta_{1,1}\boldsymbol{X_{1,1}} + \ldots + \beta_{k,1}\boldsymbol{X_{k,1}} \\
               \beta_{0,2}\mathbf{1} + \beta_{1,2}\boldsymbol{X_{1,2}} + \ldots + \beta_{k,2}\boldsymbol{X_{k,2}}
\end{pmatrix}\\
\end{eqnarray}

\

and the phylogenetic random effects and residuals are drawn from multivariate normal distributions:

\

\begin{eqnarray}
(\mathbf{b}_1, \mathbf{b}_2) &\sim& \mbox{MVN}(0, \Omega)\\
(\mathbf{e}_1, \mathbf{e}_2) &\sim& \mbox{MVN}(0, \Sigma)\\[2mm]
\Omega &=& \Sigma^{\mathrm{\tiny phy}}\otimes A \\
\Sigma &=& \Sigma^{\mathrm{\tiny res}}\otimes I
\end{eqnarray}\


Here, $\Sigma^{\mathrm{\tiny phy}}$ is an $m\times m$ phylogenetic trait-level VCV matrix and $m$ is the number of response traits in the model,

\

\begin{eqnarray}
\Sigma^{\mathrm{\tiny phy}}
= 
\begin{pmatrix}\Sigma^{\mathrm{\tiny phy}}_{11} & \Sigma^{\mathrm{\tiny phy}}_{12} & \ldots & \Sigma^{\mathrm{\tiny phy}}_{1m} \\ 
                \Sigma^{\mathrm{\tiny phy}}_{21} & \Sigma^{\mathrm{\tiny phy}}_{22} & \ldots & \Sigma^{\mathrm{\tiny phy}}_{2m} \\
                \vdots & \vdots & \ddots & \vdots \\ 
                \Sigma^{\mathrm{\tiny phy}}_{m1} & \Sigma^{\mathrm{\tiny phy}}_{m2} & \ldots & \Sigma^{\mathrm{\tiny phy}}_{mm} \\ \end{pmatrix}
&=& 
\begin{pmatrix}(\sigma^{\mathrm{\tiny phy}}_1)^2 & \rho^{\mathrm{\tiny phy}}_{12}\sigma^{\mathrm{\tiny phy}}_1 \sigma^{\mathrm{\tiny phy}}_2 & 
                \ldots & \rho^{\mathrm{\tiny phy}}_{1m}\sigma^{\mathrm{\tiny phy}}_1 \sigma^{\mathrm{\tiny phy}}_m \\ 
                \rho^{\mathrm{\tiny phy}}_{21}\sigma^{\mathrm{\tiny phy}}_2 \sigma^{\mathrm{\tiny phy}}_1 & (\sigma^{\mathrm{\tiny phy}}_2)^2 & 
                \ldots & \rho^{\mathrm{\tiny phy}}_{2m}\sigma^{\mathrm{\tiny phy}}_2 \sigma^{\mathrm{\tiny phy}}_m \\
                \vdots & \vdots & \ddots & \vdots \\ 
                \rho^{\mathrm{\tiny phy}}_{m1}\sigma^{\mathrm{\tiny phy}}_m \sigma^{\mathrm{\tiny phy}}_1 & 
                \rho^{\mathrm{\tiny phy}}_{m2}\sigma^{\mathrm{\tiny phy}}_m\sigma^{\mathrm{\tiny phy}}_2 &
                \ldots & 
                (\sigma^{\mathrm{\tiny phy}}_m)^2 \\ \end{pmatrix}\\[2mm]

\end{eqnarray}
\

such that elements of $\Sigma^{\mathrm{\tiny phy}}$ have the general form,

\

\begin{eqnarray}\
\Sigma^{\mathrm{\tiny phy}}_{ij} &=&  \rho^{\mathrm{\tiny phy}}_{ij}\sigma^{\mathrm{\tiny phy}}_i \sigma^{\mathrm{\tiny phy}}_j \\
\end{eqnarray}

\

where $\rho_{ij}$ is the correlation between traits $i$ and $j$  and $\sigma_i$ is the standard deviation of trait $i$. 

\

#### Example

To cement these concepts, as well as make clearer the estimation and interpretation of individual variance components, it is useful to explore a fully-worked, low-dimensional example. Consider the following 5-taxon phylogeny:

\

```{r}
t.toy2 <- read.tree(text='((1:1, 2:1):3, (3:3, (4:2, 5:2):1):1);')
plot(t.toy2,mar=c(5.1,0.2,0.2,0.2))
axis(1, at = seq(0,4,by=1), labels = rev(seq(0,4,by=1)), line = 0.5)
title(xlab="distance")
```

with inverse distance matrix (elements expressing the sum of shared branch lengths between two taxa, $i$ and $j$), $A$:

\

\begin{eqnarray}
A &=& 
\begin{pmatrix}A_{11} & A_{12} & A_{13} & A_{14} & A_{15} \\ 
                A_{21} & A_{22} & A_{23} & A_{24} & A_{25} \\
                A_{31} & A_{32} & A_{33} & A_{34} & A_{35} \\
                A_{41} & A_{42} & A_{43} & A_{44} & A_{45} \\
                A_{51} & A_{52} & A_{53} & A_{54} & A_{55} \\
                \end{pmatrix}
&=&                
\begin{pmatrix} 4 & 3 & 0 & 0 & 0 \\ 
                3 & 4 & 0 & 0 & 0 \\
                0 & 0 & 4 & 1 & 1 \\
                0 & 0 & 1 & 4 & 2 \\
                0 & 0 & 1 & 2 & 4 \\
                \end{pmatrix}
&=&
\begin{pmatrix} &&&&\\&&&&\\&&A_{ij}&&\\&&&&\\&&&& \end{pmatrix} \\
                
\end{eqnarray}\


For a model with two Gaussian response traits (i.e., $m = 2$), the VCV matrix for the phylogenetic random effects, $\Omega$, takes the form:

\

<!-- Can't get parenthesis  around smallmatrix. Need \mathtools but how to load?  -->

\begin{eqnarray}

\Omega
&=& 
\Sigma^{\mathrm{\tiny phy}} \otimes A
&=& 
\begin{pmatrix} \Sigma^{\mathrm{\tiny phy}}_{11}A & \Sigma^{\mathrm{\tiny phy}}_{12}A  \\ 
                \Sigma^{\mathrm{\tiny phy}}_{21}A & \Sigma^{\mathrm{\tiny phy}}_{22}A  \\
\end{pmatrix}
&=&
\begin{pmatrix} \Sigma^{\mathrm{\tiny phy}}_{11}
                \left(\begin{smallmatrix} 
                4 & 3 & 0 & 0 & 0 \\ 
                3 & 4 & 0 & 0 & 0 \\
                0 & 0 & 4 & 1 & 1 \\
                0 & 0 & 1 & 4 & 2 \\
                0 & 0 & 1 & 2 & 4 \\
                \end{smallmatrix}\right) &
                \Sigma^{\mathrm{\tiny phy}}_{12}\begin{pmatrix} &&\\&A_{ij}&\\&& \end{pmatrix} \\
                \Sigma^{\mathrm{\tiny phy}}_{21}\begin{pmatrix} &&\\&A_{ij}&\\&& \end{pmatrix} &
                \Sigma^{\mathrm{\tiny phy}}_{22}\begin{pmatrix} &&\\&A_{ij}&\\&& \end{pmatrix} \\



\end{pmatrix}

\end{eqnarray}\

\

Similarly, the VCV matrix for the residual errors, $\Sigma$, is:

\


<!-- Can't get parenthesis  around smallmatrix. Need \mathtools but how to load?  -->

\begin{eqnarray}

\Sigma
&=& 
\Sigma^{\mathrm{\tiny res}} \otimes I
&=& 
\begin{pmatrix} \Sigma^{\mathrm{\tiny res}}_{11}I & \Sigma^{\mathrm{\tiny res}}_{12}I  \\ 
                \Sigma^{\mathrm{\tiny res}}_{21}I & \Sigma^{\mathrm{\tiny res}}_{22}I  \\
\end{pmatrix}
&=&
\begin{pmatrix} \Sigma^{\mathrm{\tiny res}}_{11}
                \left(\begin{smallmatrix} 
                1 & 0 & 0 & 0 & 0 \\ 
                0 & 1 & 0 & 0 & 0 \\
                0 & 0 & 1 & 0 & 0 \\
                0 & 0 & 0 & 1 & 0 \\
                0 & 0 & 0 & 0 & 1 \\
                \end{smallmatrix}\right) &
                \Sigma^{\mathrm{\tiny res}}_{12}\begin{pmatrix} &&\\&I_{ij}&\\&& \end{pmatrix} \\
                \Sigma^{\mathrm{\tiny res}}_{21}\begin{pmatrix} &&\\&I_{ij}&\\&& \end{pmatrix} &
                \Sigma^{\mathrm{\tiny res}}_{22}\begin{pmatrix} &&\\&I_{ij}&\\&& \end{pmatrix} \\
\end{pmatrix}

\end{eqnarray}\


__N.B. This example is purely for illustrative purposes. The parameters of this model are not estimable with such a small sample size (n = 5 taxa). However, such low dimensions of $A$ and $I$ allow us to more easily and explicitly demonstrate the working of matrix manipulations.__

\

A MV model with two Gaussian response traits, such as the example provided above, can be implemented in MCMCglmm and brms as follows:

```{r eval=FALSE}

## MCMCglmm 
p2 <- list(G = list(G1 = list(V = diag(1,2), nu = 1.002)), 
           R = list(V = diag(1,2), nu = 1.002))
m.2<-MCMCglmm(cbind(y1, y2) ~ trait-1,
                random = ~us(trait):animal,
                rcov = ~us(trait):units,
                pedigree=t.toy,
                family = c("gaussian","gaussian"), 
                data = d.toy, prior=p2, 
                nitt=210000, burnin=10000, thin=200,
                pr=TRUE,verbose = FALSE)

## BRMS
b.2 <- brm(mvbind(y1, y2) ~ (1|p|gr(animal, cov = A)), 
              data = d.toy, 
              family = gaussian(), 
              data2 = list(A = A.mat),
              cores=4, chains=4, 
              iter = 6000, thin = 3
)

```


\

Explanation and comparison of syntax for MCMCglmm and brms

\

### Model 3 - Multivariate non-Gaussian

\

#### Model Explanation

The covariances between traits (the trait-level covariances) are modeled on the link scale. Gaussian responses are drawn directly from the MV structure, as a result of the identity link function. For non-Gaussian responses, the link is non trivial. E.g., For Poisson regression the mean is modeled on the log link and for the binomial, the probability of success is modeled on the logit link.

\

\begin{eqnarray}
\begin{pmatrix}\mathbf{y_1} \\ \mathbf{y_2} \\ \mathrm{log}(\boldsymbol{\lambda_3}) \\ \mathrm{logit}(\mathbf{p_4}) \end{pmatrix} &=& 
\begin{pmatrix}\boldsymbol{\mu}_1 + \mathbf{b}_1 + \mathbf{e}_1 \\
               \boldsymbol{\mu}_2 + \mathbf{b}_2 + \mathbf{e}_2 \\
               \boldsymbol{\mu}_3 + \mathbf{b}_3 + \mathbf{e}_3 \\
               \boldsymbol{\mu}_4 + \mathbf{b}_4 + \mathbf{e}_4
\end{pmatrix}\\[2mm]
\\


\mathbf{y_3} &\sim& \mathrm{Poisson}(\boldsymbol{\lambda_3}) \\
\mathbf{y_4} &\sim& \mathrm{binomial}(\mathbf{p_4})

\end{eqnarray}

\

Implementations of MV non-Gaussian models: 
```{r eval=FALSE}
# MCMCglmm

# NOTE - NEED TO USE DIFFERENT PRIORS FOR NON-GAUSSIAN RESPONSES
p3 <- list(G = list(G1 = list(V = diag(4), nu = 3.002)),
           R = list(V = diag(4), nu = 3.002))

m.3 <- MCMCglmm(cbind(y1, y2, y3, y4) ~ trait-1,
                random = ~us(trait):animal,
                rcov = ~us(trait):units,
                pedigree=t.toy,
                family = c("gaussian","gaussian","Poisson", "categorical"), 
                data = d.toy, prior=p3, 
                nitt=210000, burnin=10000, thin=200,
                pr=TRUE,verbose = FALSE)


# BRMS
bf_y1 <- bf(y1 ~ 1 + (1|p|gr(animal, cov = A)) + (1|q|obs), sigma = 0.01) + gaussian()
bf_y2 <- bf(y2 ~ 1 + (1|p|gr(animal, cov = A)) + (1|q|obs), sigma = 0.01) + gaussian()
bf_y3 <- bf(y3 ~ 1 + (1|p|gr(animal, cov = A)) + (1|q|obs)) + poisson()
bf_y4 <- bf(y4 ~ 1 + (1|p|gr(animal, cov = A)) + (1|q|obs)) + bernoulli()

b.3 <- brm(bf_y1 + bf_y2 + bf_y3 + bf_y4 + set_rescor(FALSE),
          data = d.toy, 
          family = gaussian(), 
          data2 = list(A = A.mat),
          cores=4,
          chains=4, iter = 2000, thin = 1
)

```

\
\

### Simulating data under each model

First we will simulate a pure-birth tree with 50 tips from which to generate our phylogenetic VCV matrix $A$. We will also plot the tree:

```{r}
t.toy <- sim.bdtree(b=1, d=0, stop="taxa", n=50, extinct=FALSE) # may be best not to use pure birth trees (Adams and Collyer 2018)
# code to calc VCV and take inverse 
plot.phylo(t.toy, show.tip.label=F)

```

R code showing simulations

\
\

#### NOTES

we now have three levels of (co)variance to consider; trait-level, taxon-level and residual, each of which can be estimated with appropriately structured VCV matrices.

For maths use dollars signs to get latex commands: $\lambda$

For a reference use at symbol @Royle2004 , see the bib_phmm.bib file for standard bibtex. 

\

For a sample of 5 species means:

\begin{eqnarray}
I &=&
\begin{pmatrix} 
                1 & 0 & 0 & 0 & 0 \\ 
                0 & 1 & 0 & 0 & 0 \\
                0 & 0 & 1 & 0 & 0 \\
                0 & 0 & 0 & 1 & 0 \\
                0 & 0 & 0 & 0 & 1 \\
                \end{pmatrix}
\end{eqnarray}\


#### ALTERNATIVE NOTATION

\begin{eqnarray}
\Sigma^{\mbox{phy}}_{11} = \sigma^{2,\mbox{phy}}_{1} \\
\Sigma^{phy} = A \otimes \Omega^{phy}  \\
\Sigma^{phy}_{mn} = A_{ij}\Omega^{phy}_{kl} \\
\Omega^{phy}_{kl} = \rho^{phy}_{kl}\sigma^{2,phy}_k\sigma^{2,phy}_l \\
\\
(\Sigma_p)_{11} = (\sigma_p)^2_{1} \\
\Sigma_p = A \otimes \Omega_p  \\
(\Sigma_p)_{mn} = A_{ij}(\Omega_p)_{kl} \\
(\Omega_{p})_{kl} = (\rho_{p})_{kl}(\sigma_{p})^2_k(\sigma_{p})^2_l
\end{eqnarray}\



\

# References

