---
title: "A Primer in MV-PMM"
author: "Ben Halliwell, Luke Yates & Barbara Holland"
header-includes:
  - \usepackage{mathtools}
date: "XX/XX/2022"
bibliography: bib_phmm.bib
link-citations: true
output: 
  html_document:
    number_sections: false
    toc: true 
    toc_float:
      collapsed: false
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse); library(ggplot2); library(knitr); library(brms)
library(rstan); library(geiger);library(ape);library(ape)
knitr::opts_chunk$set(echo = TRUE, dev="png",
               dpi=96)
```


### Introduction

This tutorial is associated with the article "A Primer in Multivariate Phylogenetic Mixed Modelling"

EcoEvoRxiv preprint DOI XXXXXXX

by

Ben Halliwell, Luke Yates & Barbara Holland

In this tutorial we will cover how to implement phylogenetic mixed models (PMM) in two popular R packages, MCMCglmm and brms. PMM is motivated by the realisation that, because closely related species tend to resemble each other phenotypically (a phenomenon known as phylogenetic signal), we cannot assume that species represent independent data points with respect to evolutionary hypotheses. Instead, we must test for non-independence  and weight our analyses accordingly. A central assumption of PMM therefore, is that we can combine a molecular phylogeny with a model of evolution (nominally, Brownian Motion (BM)) to generate a covariance matrix that defines the expected phenotypic similarity among taxa. The strength of this signal is optimised from the data during model fit, scaling the magnitude of effects to produce a phylogenetically informed regression. For univariate (UV) cases, this approach only estimates phylogenetic signal in the response variable, while any predictor variables in the model are assumed not to display phylogenetic signal. This assumption is commonly violated in analyses of real world data sets, where predictor variables often represent species traits that are themselves subject to phylogenetic effects. This problem is effectively addressed by moving offending predictor variables to the LHS of the model equation in multivariate (MV) implementations. This way, phylogenetic signal is simultaneously evaluated for all relevant traits, allowing (co)variances between response variables to be partitioned across multiple levels in the model hierarchy (i.e. phylogenetic and residual). This partitioning is important because phenotypic covariances (i.e., covariance on the level of traits measured), Cov(y1,y2), represent the sum of covariances on hierarchically lower levels, which need not align in magnitude or even direction. Thus, for many comparative analyses of species traits, MV-PMM should be preferred, as it offers more informative outputs with respect to evolutionary hypotheses.

While frequentist implementations of the PMM are available (e.g., pglmm() in the "phyr" package), we advocate for a Bayesian approach due to 1) greater flexibility in fitting non-Gaussian response traits 2) superiority in estimating variance components associated with hierarchical random effects; 3)....?

Throughout this tutorial, we purposefully explore a low dimensional example (tree with only 5 taxa) in order to show full workings of the matrix manipulations that underlie the models specified. This is for the sake of exposition only; analyses of the form presented will typically require much higher phylogenetic replication to produce reliable estimates of, e.g. phylogenetic correlations (Housworth et al. 2004). Thus, while mathematical workings follow a 5 taxa example, model outputs are based on analyses from simulations including 250 taxa.

\

### Model 1 - Univariate Gaussian

\

#### Model Explanation

For the case of a single Gaussian trait, our response variable, $y$, is modeled directly as a linear combination of the trait 
intercept, $\beta_0$, a phylogenetic random effect, $b$, and residual variance, $\mathbf{e}$.

\

\begin{eqnarray}
\mathbf{y} &=& \boldsymbol{\mu} + \mathbf{b} + \mathbf{e}\\[2mm]
\end{eqnarray}\


In order to capture the expected influence of co-ancestry on values $y_i$, (i.e., more similar values of $y$ among more closely related taxa), the distribution of our random effect, $b$, is multivariate normal, with mean 0 and (co)variance given by $\sigma_{phy}A$, 

\

\begin{eqnarray}
b &\sim& \mbox{MVN}(0, \sigma_{phy}A)\\
\end{eqnarray}\

<!-- The phylogenetic variance (e.g., Brownian rate), $\sigma_{phy}$, is a scalar to be estimated by the model and $A$ is a fixed $n\times n$ -->
<!-- phylogenetic VCV matrix, where $n$ is the number of taxa in the tree. This model therefore takes as input a phylogenetic tree (topology and branch lengths), which is assumed to be known without error. The $A$ matrix itself is derived by taking the inverse of the distance matrix of the tree. It gives the expected (co)variance among taxa in the value of $y$ assuming some model of evolution (e.g. BM) along the branches of that tree. -->

Here, $A$ is a fixed $n\times n$ phylogenetic VCV matrix, where $n$ is the number of taxa in the tree, and $\sigma_{phy}$, the phylogenetic variance (e.g., Brownian rate), is a scalar to be estimated by the model. This model therefore takes as input a phylogenetic tree (topology and branch lengths), which is assumed to be known without error. Technically, the $A$ matrix is derived by taking the inverse of the distance matrix of the tree. This gives the expected (co)variance among taxa in values $y$ assuming BM evolution on that tree.

\

\begin{eqnarray}
A &=& 
\begin{pmatrix}A_{11} & A_{12} & \ldots & A_{1n} \\ 
                A_{21} & A_{22} & \ldots & A_{2n} \\
                \vdots & \vdots & \ddots & \vdots \\ 
                A_{n1} & A_{n2} & \ldots & A_{nn} \\ 
\end{pmatrix}
\end{eqnarray}\


Our observation-level residual error, $\mathbf{e}$, on the other hand, is assumed to be independent of species identity. This is specified by use of the identity matrix, $I$ (i.e., an $n\times n$ matrix with 1 in all diagonal elements and 0 in all off-diagonal elements) in the distributional statement:

\

\begin{eqnarray}
\mathbf{e} &\sim& \mbox{MVN}(0, \sigma_{res}I)\\
\end{eqnarray}\

#### Example
To cement these concepts, it is useful to explore a fully-worked, low-dimensional example. Consider the following 5-taxon phylogeny with genetic distance, or dissimilarity, shown on the x axis:

\

```{r}
t.toy2 <- read.tree(text='((1:1, 2:1):3, (3:3, (4:2, 5:2):1):1);')
plot(t.toy2,mar=c(5.1,0.2,0.2,0.2))
axis(1, at = seq(0,4,by=1), labels = rev(seq(0,4,by=1)), line = 0.5)
title(xlab="distance")
```

For this phylogeny, the inverse distance matrix (elements expressing the sum of shared branch lengths between two taxa, $i$ and $j$), $A$, is:

\

\begin{eqnarray}
A &=& 
\begin{pmatrix}A_{11} & A_{12} & A_{13} & A_{14} & A_{15} \\ 
                A_{21} & A_{22} & A_{23} & A_{24} & A_{25} \\
                A_{31} & A_{32} & A_{33} & A_{34} & A_{35} \\
                A_{41} & A_{42} & A_{43} & A_{44} & A_{45} \\
                A_{51} & A_{52} & A_{53} & A_{54} & A_{55} \\
                \end{pmatrix}
&=&                
\begin{pmatrix} 4 & 3 & 0 & 0 & 0 \\ 
                3 & 4 & 0 & 0 & 0 \\
                0 & 0 & 4 & 1 & 1 \\
                0 & 0 & 1 & 4 & 2 \\
                0 & 0 & 1 & 2 & 4 \\
                \end{pmatrix}
&=&
\begin{pmatrix} &&&&\\&&&&\\&&A_{ij}&&\\&&&&\\&&&& \end{pmatrix} \\
                
\end{eqnarray}\


For numerical reasons, $A$ is often re-scaled to a correlation matrix prior to analyses, such that,

\

\begin{eqnarray}
A_{cor} &=& 
\begin{pmatrix} 1 & 0.75 & 0 & 0 & 0 \\ 
                0.75 & 1 & 0 & 0 & 0 \\
                0 & 0 & 1 & 0.25 & 0.25 \\
                0 & 0 & 0.25 & 1 & 0.5 \\
                0 & 0 & 0.25 & 0.5 & 1 \\
                \end{pmatrix}
\end{eqnarray}\



For this example, the identity matrix $I$ is:

\begin{eqnarray}
I &=& 
\begin{pmatrix} 1 & 0 & 0 & 0 & 0 \\ 
                0 & 1 & 0 & 0 & 0 \\
                0 & 0 & 1 & 0 & 0 \\
                0 & 0 & 0 & 1 & 0 \\
                0 & 0 & 0 & 0 & 1 \\
                \end{pmatrix}
\end{eqnarray}\

\

#### Model Implementation

Implementing these models in R is fairly straight forward in both MCMCglmm and brms. But first, let's simulate some data.
To simulate data under this model, we simply construct our variable, $y$, from the linear combination of predictors specified by
the model, by supplying vectors of length $n$ for each of the quantities $\beta_0$, $b$, and $\mathbf{e}$. For Gaussian variables
this can be done directly, while for non-Gaussian variables we must make predictions on the link scale (see section 3):

\

```{r eval=FALSE}

## SIMULATION

# simulate tree
n = 250 # number of species
t.toy <- geiger::sim.bdtree(b=1, d=0, stop="taxa", n=n, extinct=FALSE) # N.B. may be best not to use pure birth trees (Adams and Collyer 2018)
t.toy <- multi2di(t.toy) # ensure tree is fully bifurcating

# create VCV matrix from tree, scale to correlation matrix with corr = T
A.mat <- ape::vcv.phylo(t.toy, corr = T)

# phylogenetic VCV matrix
A = A.mat

# identity matrix
I = diag(n)

# intercept
beta = 1

# phylogenetic variance
sigma_phy = 0.5

# residual variance
sigma_res = 0.1

# simulate phylogenetic random effects as one draw from a MVN
b <- MASS::mvrnorm(1,rep(0,n),sigma_phy*A)

# simulate residuals
e <- MASS::mvrnorm(1,rep(0,n),sigma_res*I)

# construct response traits from the linear predictor
y <- beta + b + e

# generate df
animal <- t.toy$tip.label # "animal" is a reserved term in MCMCglmm for taxon ID
d.toy <- data.frame(animal,y)

## PLOTS TO EXPLORE THIS SIMULATED DATA?

```

\

With our tree and trait data simulated, we can now fit the model:

\

```{r eval=FALSE}

## MCMCglmm 
p.m.1 <- list(G = list(G1 = list(V = 1, nu = 0.002)), 
              R = list(V = 1, nu = 0.002))
m.1 <- MCMCglmm(y ~ 1,
                random = ~animal,
                pedigree = t.toy, 
                family = c("gaussian"), 
                data = d.toy, 
                prior = p.m.1,
                nitt = 210000, burnin = 10000, thin = 200,
                pr = TRUE, verbose = FALSE)


## BRMS
p.b.1 <- list()
b.1 <- brm(y ~ 1 + (1|gr(animal, cov = A)), 
            data = d.toy,
            prior = p.b.1
            family = gaussian(), 
            data2 = list(A = A.mat),
            cores = 4, chains = 4, iter = 6000, thin = 3)

```

\

#### Model Syntax (MCMCglmm vs. brms) 

The code above fits the same model in MCMCglmm and brms. In MCMCglmm, `animal` is a reserved term used to identify individuals/taxa in a quantitative genetic/phylogenetic analysis. The random effect specification `~animal` instructs MCMCglmm to fit a random effect at the individual/taxon level with covariance structure supplied via the `pedigree` argument. For `pedigree`, MCMCglmm takes the tree directly and performs the necessary steps to derive the inverse distance matrix $A$ under the hood. The remaining code specifies the sampler settings, whether the post. dist. of random effect should be saved and whether to print sampler progress to the console.



Note that it is not necessary to specify a prior to fit this model in either MCMCglmm or brms; if no prior is specified, default priors are used automatically. However, we choose to explicitly specify (currently default) priors for each package to ensure repeatability, should the default priors of either package change in the future. Furthermore, a primary differentiating factor between these two packages is that MCMCglmm uses Gibbs sampling, where-as brms uses Hamiltonian sampling. These different computational approaches to MCMC sampling place different requirements on prior specification, which affects the user experience.

\

#### Diangostics

\

#### Interpretation

If we inspect the summaries for each model fit...

It can be seen from (X; Halliwell, Yates & Holland 2022) that the marginal Gaussian UV-PMM is mathematically equivalent to PGLS.

\

### Model 2 - Multivariate Gaussian

\

#### Model Explanation

In a multivariate (MV) implementation, our response variables are modelled jointly, allowing both phylogenetic and residual (co)variances to be estimated. With this design, the phylogenetic random effects must also be realised jointly as a multivariate normal, with variance given by $\Omega$, the Kronecker product of a trait-level correlation matrix $\Sigma^{\mathrm{\tiny phy}}$ and the phylogenetic VCV matrix, $A$. With $m$ traits (response variables) and $n$ taxa in the phylogeny, $\Omega$ is an $mn\times mn$ VCV matrix containing phylogenetic (co)variances for all traits as well as between all traits. Notice that the two taxon-level variance structures we encountered in the univariate case above ($A$ and $I$) remain, but we must now also consider trait-level covariance operating at both the phylogenetic and residual level. This is achieved via the Kronecker operation, e.g. $\Omega = \Sigma^{\mathrm{\tiny phy}}\otimes A$. A MV-PMM with two Gaussian response variables takes the form,


\


\begin{eqnarray}
\begin{pmatrix}\mathbf{y_1} \\
\mathbf{y_2} \end{pmatrix} &=& 
\begin{pmatrix}\boldsymbol{\mu}_1 + \mathbf{b}_1 + \mathbf{e}_1 \\
               \boldsymbol{\mu}_2 + \mathbf{b}_2 + \mathbf{e}_2
\end{pmatrix}\\
\end{eqnarray}

\

where fixed effects are expressed as a linear combination of predictors,

\

\begin{eqnarray}
\begin{pmatrix}\boldsymbol{\mu_1} \\ \boldsymbol{\mu_2} \end{pmatrix} &=& 
\begin{pmatrix}\beta_{0,1}\mathbf{1} + \beta_{1,1}\boldsymbol{X_{1,1}} + \ldots + \beta_{k,1}\boldsymbol{X_{k,1}} \\
               \beta_{0,2}\mathbf{1} + \beta_{1,2}\boldsymbol{X_{1,2}} + \ldots + \beta_{k,2}\boldsymbol{X_{k,2}}
\end{pmatrix}\\
\end{eqnarray}

\

and the phylogenetic random effects and residuals are drawn from multivariate normal distributions,

\

\begin{eqnarray}
(\mathbf{b}_1, \mathbf{b}_2) &\sim& \mbox{MVN}(0, \Omega)\\
(\mathbf{e}_1, \mathbf{e}_2) &\sim& \mbox{MVN}(0, \Sigma)\\[2mm]
\Omega &=& \Sigma^{\mathrm{\tiny phy}}\otimes A \\
\Sigma &=& \Sigma^{\mathrm{\tiny res}}\otimes I
\end{eqnarray}\


Here, $\Sigma^{\mathrm{\tiny phy}}$ is an $m\times m$ phylogenetic trait-level VCV matrix and $m$ is the number of response traits in the model,

\

\begin{eqnarray}
\Sigma^{\mathrm{\tiny phy}}
= 
\begin{pmatrix}\Sigma^{\mathrm{\tiny phy}}_{11} & \Sigma^{\mathrm{\tiny phy}}_{12} & \ldots & \Sigma^{\mathrm{\tiny phy}}_{1m} \\ 
                \Sigma^{\mathrm{\tiny phy}}_{21} & \Sigma^{\mathrm{\tiny phy}}_{22} & \ldots & \Sigma^{\mathrm{\tiny phy}}_{2m} \\
                \vdots & \vdots & \ddots & \vdots \\ 
                \Sigma^{\mathrm{\tiny phy}}_{m1} & \Sigma^{\mathrm{\tiny phy}}_{m2} & \ldots & \Sigma^{\mathrm{\tiny phy}}_{mm} \\ \end{pmatrix}
&=& 
\begin{pmatrix}(\sigma^{\mathrm{\tiny phy}}_1)^2 & \rho^{\mathrm{\tiny phy}}_{12}\sigma^{\mathrm{\tiny phy}}_1 \sigma^{\mathrm{\tiny phy}}_2 & 
                \ldots & \rho^{\mathrm{\tiny phy}}_{1m}\sigma^{\mathrm{\tiny phy}}_1 \sigma^{\mathrm{\tiny phy}}_m \\ 
                \rho^{\mathrm{\tiny phy}}_{21}\sigma^{\mathrm{\tiny phy}}_2 \sigma^{\mathrm{\tiny phy}}_1 & (\sigma^{\mathrm{\tiny phy}}_2)^2 & 
                \ldots & \rho^{\mathrm{\tiny phy}}_{2m}\sigma^{\mathrm{\tiny phy}}_2 \sigma^{\mathrm{\tiny phy}}_m \\
                \vdots & \vdots & \ddots & \vdots \\ 
                \rho^{\mathrm{\tiny phy}}_{m1}\sigma^{\mathrm{\tiny phy}}_m \sigma^{\mathrm{\tiny phy}}_1 & 
                \rho^{\mathrm{\tiny phy}}_{m2}\sigma^{\mathrm{\tiny phy}}_m\sigma^{\mathrm{\tiny phy}}_2 &
                \ldots & 
                (\sigma^{\mathrm{\tiny phy}}_m)^2 \\ \end{pmatrix}\\[2mm]

\end{eqnarray}
\

such that elements of $\Sigma^{\mathrm{\tiny phy}}$ have the general form,

\

\begin{eqnarray}\
\Sigma^{\mathrm{\tiny phy}}_{ij} &=&  \rho^{\mathrm{\tiny phy}}_{ij}\sigma^{\mathrm{\tiny phy}}_i \sigma^{\mathrm{\tiny phy}}_j \\
\end{eqnarray}

\

where $\rho_{ij}$ is the correlation between traits $i$ and $j$  and $\sigma_i$ is the standard deviation of trait $i$. Thus, returning to our 5-taxon example, a MV model with two Gaussian response traits (i.e., $m = 2$) will have VCV matrix for the phylogenetic random effects, $\Omega$, of the form:

\

<!-- Can't get parenthesis  around smallmatrix. Need \mathtools but how to load?  -->

\begin{eqnarray}

\Omega
&=& 
\Sigma^{\mathrm{\tiny phy}} \otimes A
&=& 
\begin{pmatrix} \Sigma^{\mathrm{\tiny phy}}_{11}A & \Sigma^{\mathrm{\tiny phy}}_{12}A  \\ 
                \Sigma^{\mathrm{\tiny phy}}_{21}A & \Sigma^{\mathrm{\tiny phy}}_{22}A  \\
\end{pmatrix}
&=&
\begin{pmatrix} \Sigma^{\mathrm{\tiny phy}}_{11}
                \left(\begin{smallmatrix} 
                4 & 3 & 0 & 0 & 0 \\ 
                3 & 4 & 0 & 0 & 0 \\
                0 & 0 & 4 & 1 & 1 \\
                0 & 0 & 1 & 4 & 2 \\
                0 & 0 & 1 & 2 & 4 \\
                \end{smallmatrix}\right) &
                \Sigma^{\mathrm{\tiny phy}}_{12}\begin{pmatrix} &&\\&A_{ij}&\\&& \end{pmatrix} \\
                \Sigma^{\mathrm{\tiny phy}}_{21}\begin{pmatrix} &&\\&A_{ij}&\\&& \end{pmatrix} &
                \Sigma^{\mathrm{\tiny phy}}_{22}\begin{pmatrix} &&\\&A_{ij}&\\&& \end{pmatrix} \\



\end{pmatrix}

\end{eqnarray}\

\

Similarly, the VCV matrix for the residual errors, $\Sigma$, is:

\


<!-- Can't get parenthesis  around smallmatrix. Need \mathtools but how to load?  -->

\begin{eqnarray}

\Sigma
&=& 
\Sigma^{\mathrm{\tiny res}} \otimes I
&=& 
\begin{pmatrix} \Sigma^{\mathrm{\tiny res}}_{11}I & \Sigma^{\mathrm{\tiny res}}_{12}I  \\ 
                \Sigma^{\mathrm{\tiny res}}_{21}I & \Sigma^{\mathrm{\tiny res}}_{22}I  \\
\end{pmatrix}
&=&
\begin{pmatrix} \Sigma^{\mathrm{\tiny res}}_{11}
                \left(\begin{smallmatrix} 
                1 & 0 & 0 & 0 & 0 \\ 
                0 & 1 & 0 & 0 & 0 \\
                0 & 0 & 1 & 0 & 0 \\
                0 & 0 & 0 & 1 & 0 \\
                0 & 0 & 0 & 0 & 1 \\
                \end{smallmatrix}\right) &
                \Sigma^{\mathrm{\tiny res}}_{12}\begin{pmatrix} &&\\&I_{ij}&\\&& \end{pmatrix} \\
                \Sigma^{\mathrm{\tiny res}}_{21}\begin{pmatrix} &&\\&I_{ij}&\\&& \end{pmatrix} &
                \Sigma^{\mathrm{\tiny res}}_{22}\begin{pmatrix} &&\\&I_{ij}&\\&& \end{pmatrix} \\
\end{pmatrix}

\end{eqnarray}\


__N.B. This example is purely for illustrative purposes. The parameters of this model are not estimable with such low phylogenetic replication (n = 5 taxa). However, the low dimension of $A$ and $I$ allow us to explicitly show the working of matrix manipulations.__

\

#### Model Implementation

A MV model with two Gaussian response traits, such as the example provided above, can be implemented in MCMCglmm and brms as follows:

```{r eval=FALSE}

## MCMCglmm 
p.m.2 <- list(G = list(G1 = list(V = diag(2), nu = 1.002)), 
              R = list(V = diag(2), nu = 1.002))
m.2 <- MCMCglmm(cbind(y1, y2) ~ trait-1,
                random = ~us(trait):animal,
                rcov = ~us(trait):units,
                pedigree = t.toy,
                family = c("gaussian","gaussian"), 
                data = d.toy, prior=p.m.2, 
                nitt=210000, burnin=10000, thin=200,
                pr = TRUE, verbose = FALSE)

## BRMS
p.b.2 <- ()
b.2 <- brm(mvbind(y1, y2) ~ (1|a|gr(animal, cov = A)), # the index |a| specified to estimate the correlation between y1 and y2 at the animal group level
            data = d.toy,
            prior = p.b.2,
            family = gaussian(), 
            data2 = list(A = A.mat),
            cores = 4, chains = 4, iter = 6000, thin = 3)

```


\

Explanation and comparison of syntax for MCMCglmm and brms

\

### Model 3 - Multivariate Non-Gaussian (Gaussian, Bernoulli)

\

#### Model Explanation

In previous simulations, we saw that Gaussian responses can be drawn directly from the MV structure, i.e., modeled via the identity link function. For non-Gaussian variables, (co)variances must be modeled on the link scale, which may be non-trivial, e.g., for Poisson regression the mean is modeled with the log link while for binomial data, the probability of success is modeled with the logit link.

\

\begin{eqnarray}
\begin{pmatrix}\mathbf{y_1} \\ \mathbf{y_2} \\ \mathrm{log}(\boldsymbol{\lambda_3}) \\ \mathrm{logit}(\mathbf{p_4}) \end{pmatrix} &=& 
\begin{pmatrix}\boldsymbol{\mu}_1 + \mathbf{b}_1 + \mathbf{e}_1 \\
               \boldsymbol{\mu}_2 + \mathbf{b}_2 + \mathbf{e}_2 \\
               \boldsymbol{\mu}_3 + \mathbf{b}_3 + \mathbf{e}_3 \\
               \boldsymbol{\mu}_4 + \mathbf{b}_4 + \mathbf{e}_4
\end{pmatrix}\\[2mm]
\\


\mathbf{y_3} &\sim& \mathrm{Poisson}(\boldsymbol{\lambda_3}) \\
\mathbf{y_4} &\sim& \mathrm{binomial}(\mathbf{p_4})

\end{eqnarray}

\

Implementations of MV non-Gaussian models: 
```{r eval=FALSE}
# MCMCglmm

# NOTE - NEED TO USE DIFFERENT PRIORS FOR NON-GAUSSIAN RESPONSES
p3 <- list(G = list(G1 = list(V = diag(4), nu = 3.002)),
           R = list(V = diag(4), nu = 3.002))

m.3 <- MCMCglmm(cbind(y1, y2, y3, y4) ~ trait-1,
                random = ~us(trait):animal,
                rcov = ~us(trait):units,
                pedigree=t.toy,
                family = c("gaussian","gaussian","Poisson", "categorical"), 
                data = d.toy, prior=p3, 
                nitt=210000, burnin=10000, thin=200,
                pr=TRUE,verbose = FALSE)


# BRMS
bf_y1 <- bf(y1 ~ 1 + (1|p|gr(animal, cov = A)) + (1|q|obs), sigma = 0.01) + gaussian()
bf_y2 <- bf(y2 ~ 1 + (1|p|gr(animal, cov = A)) + (1|q|obs), sigma = 0.01) + gaussian()
bf_y3 <- bf(y3 ~ 1 + (1|p|gr(animal, cov = A)) + (1|q|obs)) + poisson()
bf_y4 <- bf(y4 ~ 1 + (1|p|gr(animal, cov = A)) + (1|q|obs)) + bernoulli()

b.3 <- brm(bf_y1 + bf_y2 + bf_y3 + bf_y4 + set_rescor(FALSE),
          data = d.toy, 
          family = gaussian(), 
          data2 = list(A = A.mat),
          cores=4,
          chains=4, iter = 2000, thin = 1
)

```

\
\

### Simulating data under each model

First we will simulate a pure-birth tree with 50 tips from which to generate our phylogenetic VCV matrix $A$. We will also plot the tree:

```{r}
t.toy <- sim.bdtree(b=1, d=0, stop="taxa", n=50, extinct=FALSE) # may be best not to use pure birth trees (Adams and Collyer 2018)
# code to calc VCV and take inverse 
plot.phylo(t.toy, show.tip.label=F)

```

R code showing simulations

\
\

#### NOTES

we now have three levels of (co)variance to consider; trait-level, taxon-level and residual, each of which can be estimated with appropriately structured VCV matrices.

For maths use dollars signs to get latex commands: $\lambda$

For a reference use at symbol @Royle2004 , see the bib_phmm.bib file for standard bibtex. 

\

For a sample of 5 species means:

\begin{eqnarray}
I &=&
\begin{pmatrix} 
                1 & 0 & 0 & 0 & 0 \\ 
                0 & 1 & 0 & 0 & 0 \\
                0 & 0 & 1 & 0 & 0 \\
                0 & 0 & 0 & 1 & 0 \\
                0 & 0 & 0 & 0 & 1 \\
                \end{pmatrix}
\end{eqnarray}\


#### ALTERNATIVE NOTATION

\begin{eqnarray}
\Sigma^{\mbox{phy}}_{11} = \sigma^{2,\mbox{phy}}_{1} \\
\Sigma^{phy} = A \otimes \Omega^{phy}  \\
\Sigma^{phy}_{mn} = A_{ij}\Omega^{phy}_{kl} \\
\Omega^{phy}_{kl} = \rho^{phy}_{kl}\sigma^{2,phy}_k\sigma^{2,phy}_l \\
\\
(\Sigma_p)_{11} = (\sigma_p)^2_{1} \\
\Sigma_p = A \otimes \Omega_p  \\
(\Sigma_p)_{mn} = A_{ij}(\Omega_p)_{kl} \\
(\Omega_{p})_{kl} = (\rho_{p})_{kl}(\sigma_{p})^2_k(\sigma_{p})^2_l
\end{eqnarray}\



\

# References

